---
title: "Steve Smith"
output: rmarkdown::github_document
---

Load in packages

```{r packages}
library(readr)
library(dplyr)
library(lubridate)
library(ggplot2)
```

Read in and clean the data. Downloaded from [CricInfo](http://stats.espncricinfo.com/ci/engine/player/267192.html?class=1;template=results;type=allround;view=innings) on 2019-09-06

```{r ssdata}
ssdata <- read_csv(file = "steve_smith_stats.csv", skip = 1)

Encoding(ssdata$opposition) <- "UTF-8"

ssdata %<>%
  mutate(start_date = dmy(start_date),
         opposition = iconv(opposition, "UTF-8", "UTF-8", sub = ""),
         opposition = sub("^v", "", opposition),
         not_out = grepl("\\*$", score),
         score = sub("\\*$", "", score),
         score = as.integer(score),
         overs = ifelse(overs == "-", NA, overs),
         inningstype = ifelse(is.na(overs), "bat", "bowl"),
         conceded = as.integer(conceded),
         wickets = as.integer(wickets),
         catches = as.integer(catches),
         stumpings = as.integer(stumpings),
         oversi = as.integer(overs),
         year = year(start_date)) %>%
  select(-X12)

ssdata
```
Separate out the batting data
```{r ss_bat}
ss_bat <- ssdata %>%
  filter(inningstype == "bat") %>%
  filter(!is.na(score))

ss_bat
```
Get his batting average in each year
```{r ss_ba}
ss_ba <- ss_bat %>%
  group_by(year) %>%
  summarise(runs = sum(score),
            ninnings = n(),
            notouts = sum(not_out),
            avg = sum(score)/(n() - sum(not_out)))

ss_ba
```
And career average
```{r ss career average}
ss_bat %>%
  summarise(avg = sum(score)/(n() - sum(not_out)))
```

What does it look like if we exclude the early years...

Here's his career average excluding 2010
```{r ca ex 2010}
ss_bat %>%
  filter(year != 2010) %>%
  summarise(avg = sum(score)/(n() - sum(not_out)))
```

Excluding 2010-11
```{r ca ex 10-11}
ss_bat %>%
  filter(year != 2010) %>%
  filter(year != 2011) %>%
  summarise(avg = sum(score)/(n() - sum(not_out)))
```
And excluding up to 2013, an inconveniently lean year
```{r ca ex 10-13}
ss_bat %>%
  filter(year != 2010) %>%
  filter(year != 2011) %>%
  filter(year != 2013) %>%
  summarise(avg = sum(score)/(n() - sum(not_out)))
```
What is the overall progression of Steve Smith's yearly average?
Line is smoothed average, points are yearly average, and colour scale is number of innings that year.
```{r plot smooth}
ggplot(ss_ba) +
  geom_smooth(aes(x = year, y = avg), col = "black") +
  geom_point(aes(x = year, y = avg, col = ninnings), size = 3) +
  scale_colour_viridis_c(guide = guide_colourbar(title = "No.\ninnings")) +
  theme_classic() +
  labs(x = "Year", y = "Yearly batting average")
```

**There's clearly an unequivocal upward trend, so let's make some egregious predictions about how good his average will be at the end of his career.**

So, we fit a simple exponential model, and a linear model, and make predictions from the model fits.
```{r sslm}
sslm <- lm(avg ~ year, data = ss_ba)
ssem <- lm(log(avg) ~ year, data = ss_ba)

lin.preds <- predict(sslm, list(year = 2010:2030))
exp.preds <- exp(predict(ssem, list(year = 2010:2030)))

ss_lin_preds <- tibble(year = 2010:2030, avg = lin.preds)
ss_exp_preds <- tibble(year = 2010:2030, avg = exp.preds)
```


And what do the results show?

As above with points showing annual batting averages, black lines are Steve Smith's predicted annual average, and the red line is the Don's career average of 99.94.
```{r preds plot}
sspredplot <- ggplot() +
  geom_point(data = ss_ba, aes(x = year, y = avg, col = ninnings), size = 3) +
  geom_line(data = ss_lin_preds, aes(x = year, y = avg), size = 1) +
  geom_line(data = ss_exp_preds, aes(x = year, y = avg), size = 1) +
  geom_hline(yintercept = 99.94, col = "red") +
  scale_colour_viridis_c(guide = guide_colourbar(title = "No.\ninnings")) +
  theme_classic() +
  labs(x = "Year", y = "Yearly batting average")

sspredplot
```

This is predicts that Steve Smith will have an average in each year higher than the Don's career average every year from around 2020.

*Oh my gosh it's already higher than that this year, he's even better than these models predict.*

**Ahhhh he's going to average over 100 each year until he retires at 41 and might get as high as 250!!!**

The facts don't lie.


And here are the predicted yearly averages for the linear model
```{r ss_lin_preds}
ss_lin_preds
```
... and the exponential model
```{r ss_exp_preds}
ss_exp_preds
```

```{r}
png(res = 300, height = 1000, width = 2000)

sspredplot

dev.off()
```

